\documentclass{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\begin{document}

\title{Issue XLIV: Adjoint String of Identity Modalities}
\author{Namdak Tonpa}
\date{May 2025}

\maketitle

\begin{abstract}
In Homotopy Type Theory (HoTT), identity systems (Contractible, Strict Id, Quotient, Isomorphism, Path = Equivalence) are modeled as modalities in the $\infty$-topos $\mathcal{E} = \infty\text{Grp}$. We construct a non-degenerate adjoint quadruple extending the Jacobs-Lawvere triple $C \dashv \text{Id}_A \dashv Q(-/\sim)$, incorporating Isomorphism and Path = Equivalence. The modalities are ordered by adjointness: Contractible $\leq$ Strict Id $\leq$ Quotient $\leq$ Isomorphism $\leq$ Path = Equivalence, reflecting their structure in HoTT, where Strict Id, Quotient, and Path = Equivalence are mere propositions for h-sets, while Isomorphism is not.
\end{abstract}

\section{Introduction}
Homotopy Type Theory (HoTT) provides a framework for reasoning about equality via the identity type $\text{Id}_A(x,y)$. In the $\infty$-topos $\mathcal{E} = \infty\text{Grp}$, identity systems are modalities (monads), ordered by adjointness. The classical Jacobs-Lawvere adjunction triple $C \dashv \text{Id}_A \dashv Q(-/\sim)$ captures Contractible, Strict Id, and Quotient. We extend this to a quadruple, including Isomorphism and Path = Equivalence, respecting the HoTT equivalence of Path and Equivalence and the propositional nature of Strict Id, Quotient, and Path = Equivalence for h-sets.

\section{Identity Systems as Modalities}
\begin{definition}
In HoTT, the identity systems are:
\begin{itemize}
    \item \textbf{Contractible}: (-1)-truncated types, mere propositions.
    \item \textbf{Strict Id}: $\text{Id}_A(x,y)$ for h-sets (0-truncated), a mere proposition.
    \item \textbf{Quotient}: Set-quotients $A / \sim$, 0-truncated, equivalent to Strict Id.
    \item \textbf{Isomorphism}: $\text{iso}_A(x,y)$, a triple $(f, g, p)$, not a mere proposition.
    \item \textbf{Path = Equiv}: $\text{Id}_A(x,y) \simeq (x \simeq y)$, equivalent in HoTT.
\end{itemize}
\end{definition}

In $\mathcal{E} = \infty\text{Grp}$, we define categories:
\begin{itemize}
    \item $\mathcal{E}_{\text{contr}} = \mathcal{E}_{\leq -1}$: Mere propositions.
    \item $\mathcal{E}_{\text{strict}} = \mathcal{E}_{\leq 0} \cong \text{Set}$: h-sets (Strict Id).
    \item $\mathcal{E}_{\text{quot}} = \mathcal{E}_{\leq 0} \cong \text{Set}$: h-sets (Quotient).
    \item $\mathcal{E}_{\text{iso}} \cong \mathcal{E}$: $\infty$-groupoids with isomorphisms.
    \item $\mathcal{E}_{\text{path/equiv}} \cong \mathcal{E}$: $\infty$-groupoids with paths/equivalences.
\end{itemize}

\section{Adjoint Quadruple}
The Jacobs-Lawvere triple $C \dashv \text{Id}_A \dashv Q(-/\sim)$ is extended to a non-degenerate adjoint quadruple:

\[
\mathcal{E}_{\text{contr}} \xrightarrow{F_4} \mathcal{E}_{\text{strict}} \xrightarrow{F_3} \mathcal{E}_{\text{quot}} \xrightarrow{F_2} \mathcal{E}_{\text{iso}} \xrightarrow{F_1} \mathcal{E}_{\text{path/equiv}}
\]

\begin{theorem}
The functors form an adjoint quadruple with non-degenerate adjunctions:
\[
F_4 \dashv U_4, \quad F_3 \dashv U_3, \quad F_2 \dashv U_2, \quad F_1 \dashv U_1
\]
\end{theorem}

\begin{itemize}
    \item \textbf{$F_4: \mathcal{E}_{\text{contr}} \to \mathcal{E}_{\text{strict}}$}: Inclusion of (-1)-truncated objects into 0-truncated objects. Right adjoint $U_4$: (-1)-truncation, $U_4(X) = \|X\|_{-1}$.
    \item \textbf{$F_3: \mathcal{E}_{\text{strict}} \to \mathcal{E}_{\text{quot}}$}: Canonical map to quotient structure, viewing h-sets as quotiented by trivial relations. Right adjoint $U_3$: Inverse map preserving h-set structure.
    \item \textbf{$F_2: \mathcal{E}_{\text{quot}} \to \mathcal{E}_{\text{iso}}$}: Inclusion of h-sets into $\mathcal{E}$, $\text{core}(X) \cong X$. Right adjoint $U_2$: 0-truncation, $U_2(X) = \|X\|_0$.
    \item \textbf{$F_1: \mathcal{E}_{\text{iso}} \to \mathcal{E}_{\text{path/equiv}}$}: Canonical inclusion of $\infty$-groupoids with isomorphisms into full $\infty$-groupoids with paths/equivalences. Right adjoint $U_1$: Core map, preserving isomorphism structure.
\end{itemize}

\section{Ordering by Adjointness}
The adjunctions induce the ordering:
\[
\text{Contractible} \leq \text{Strict Id} \leq \text{Quotient} \leq \text{Isomorphism} \leq \text{Path = Equivalence}
\]
\begin{itemize}
    \item \textbf{Contractible}: Coarsest, mere propositions ((-1)-truncated).
    \item \textbf{Strict Id}: h-sets, $\text{Id}_A(x,y)$ is a mere proposition.
    \item \textbf{Quotient}: Equivalent to Strict Id, 0-truncated set-quotients.
    \item \textbf{Isomorphism}: $\text{iso}_A(x,y)$ is not a mere proposition for general types.
    \item \textbf{Path = Equivalence}: Finest, full $\infty$-groupoid structure, equivalent via univalence.
\end{itemize}

\section{Conclusion}
The adjoint quadruple extends the Jacobs-Lawvere triple, capturing the structure of identity systems in HoTT. The ordering reflects their increasing complexity, with Strict Id, Quotient, and Path = Equivalence collapsing to mere propositions for h-sets, while Isomorphism retains higher structure. Future work could explore these adjunctions in other $\infty$-topoi or specific CTT models.

\end{document}
