% copyright (c) 2018 Groupoid Infinity

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{url}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{matrix}
\usetikzlibrary{babel}
\usepackage{listings}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\lstMakeShortInline[mathescape=true,columns=fixed]|

\def\mapright#1{\xrightarrow{{#1}}}
\def\mapleft#1{\xleftarrow{{#1}}}
\def\mapup#1{\Big\uparrow\rlap{\raise2pt{\scriptstyle{#1}}}}
\def\mapupl#1{\Big\uparrow\llap{\raise2pt{\scriptstyle{#1}}}}
\def\mapdown#1{\Big\downarrow\rlap{\raise2pt{\scriptstyle{#1}}}}
\def\mapdownl#1{\Big\downarrow\llap{\raise2pt{\scriptstyle{#1}}}}
\def\mapdiagl#1{\vcenter{\searrow}\rlap{\raise2pt{\scriptstyle{#1}}}}
\def\mapdiagr#1{\vcenter{\swarrow}\rlap{\raise2pt{\scriptstyle{#1}}}}

\lstset{basicstyle=\small,inputencoding=utf8}

\begin{document}

\title{Mathematical Components}
\author{Maxim Sokhatsky}
\date{
    $^1$ National Technical University of Ukraine \\
    \small Igor Sikorsky Kyiv Polytechnical Institute\\
    \today
}

\maketitle

\begin{abstract}

Each language implementation needs to be checked. The one of possible test cases
for cubical type checkers is the direct embedding of type theory model intro the language of type checker.
As Martin-Löf Type Theory is formulated using 5 types of rules, we construct aliases
for host language primitives and use type checker to prove theorems about itself.

This could be seen as ultimate test sample for type checker as
intro-elimination fusion resides in beta-eta rules, so by proving them
we prove properties of the host type checker.

{\bf Keywords}: Cubical Type Theory, Martin-Löf Type Theory
\end{abstract}
\vspace{2cm}
\tableofcontents

\section*{Intro}

Each language implementation needs to be checked. The one of possible test cases
for cubical type checkers is the direct embedding of type theory model intro the language of type checker.
As Martin-Löf Type Theory is formulated using 5 types of rules, we construct aliases
for host language primitives and use type checker to prove theorems about itself.
This could be seen as ultimate test sample for type checker as
intro-elimination fusion resides in beta-eta rules, so by proving them
we prove properties of the host type checker.
This technique of direct embedding of the model into the type checker primitives
was also used by authors to prove that Category of Sets is Cartesian Closed.

The types used in this course is founded on top of cubical modules
each falling into one of 15 categories:

(i) MLTT Types: pi, sigma, path;
(ii) Set Theory: prop, set, ordinal, hedberg;
(iii) Run-Time Inductive Types: proto, maybe, bool, nat, list, stream, vector;
(iv) Abstract Algebra: algebra;
(v) Control Structures: control;
(vi) Recursion Schemes: recursion;
(vii) Univalent Foundations: equiv, retract, iso, univ;
(viii) Higher Inductive Types: interval, interval, s1, s2, pullback, pushout, suspension, quotient, trunc;
(ix) Process Calculus: process;
(x) Category Theory and Topos Theory: cat, fun, sip, adj, ump, cones, topos, category;
(xi) Contextual Categories: cwf, csystem;
(xii) Languages: mltt, infinity;
(xiii) Algebraic Geometry: pointed, euler, seq, hopf, cw;
(xiv) Cartan Geometry in Cohesive Topos: infinitesimal, etale, manifold, bundle, shape, sharp;
(xv) K-Theory: k\_theory, swaptrans, bishop, subtype;

\subsection*{Cubical Syntax}

The BNF notation of |cubicaltt| consists of 1) telescopes (contexts or sigma chains);
2) inductive data definitions (sum chains); 3) split eliminator;
4) branches of split eliminators; 5) pure dependent type theory syntax.
It also has where, import, module constructions.

\begin{lstlisting}[mathescape=true]
def := ${\bf data}$ id tele = sum + id tele : exp = exp +
       id tele : exp ${\bf where}$ def

exp := cotele*exp + cotele$\rightarrow$exp + exp$\rightarrow$exp + (exp) + app + id +
       (exp,exp) + \ cotele$\rightarrow$exp + ${\bf split}$ cobrs + exp${\bf.1}$ + exp${\bf.2}$
\end{lstlisting}

\begin{lstlisting}[mathescape=true]
     0 := #empty         imp    := [ ${\bf import}$ id ]
   brs := 0 + cobrs      tele   := 0 + cotele
   app := exp exp        cotele := ( exp : exp ) tele
    id := [ #nat ]       sum    := 0 + id tele + id tele | sum
   ids := [ id ]         br     := ids $\rightarrow$ exp
 codec := def dec        mod    := ${\bf module}$ id ${\bf where}$ imp dec
   dec := 0 + codec      cobrs  := | br brs
\end{lstlisting}

\newpage
\section{Martin-Löf Type Theory}

Martin-Löf Type Theory (MLTT) contains $\Pi$, $\Sigma$, Id, W, Nat, List types.
Id types were added in 1984 while original MLTT was introduced in 1972.
Predicative Universe Hierarchy was added in 1975.
Despite original MLTT contains Id types that preserve uniquness of identity
proofs (UIP), we introduce here homotopical univalent heterogeneous Path
interval types with higher equalities ($\infty$-Groupoid interpretation).
Thus, allowing to proof all the MLTT rules as a whole by using reflection rule, connections and path applications.

\subsection{Pi}

$\Pi$ is a dependent function type, the generalization of functions.
As a function it can serve the wide range of mathematical constructions,
objects, types, or spaces: sets, functions, polynomial functors, infinitesimals, $\infty$-groupoids,
topological $\infty$-groupoid, CW-complexes,
categories, languages, etc. We give here nearest isomorphism of $\Pi$-types,
the fibrations or fiber bundles. The next isomorphism of functions are functors.

\begin{definition} ($\Pi$-Formation).
\begin{lstlisting}
Pi (A: U) (P: B -> U): U = (x: A) -> B x
\end{lstlisting}
\end{definition}

\begin{definition} ($\Pi$-Introduction).
\begin{lstlisting}
lambda (A B: U) (b: B): A -> B = \ (x: A) -> b
lam (A:U) (B: A -> U) (a:A) (b:B a) : A -> B a = \ (x: A) -> b
\end{lstlisting}
\end{definition}

\begin{definition} ($\Pi$-Elimination).
\begin{lstlisting}
apply (A B: U) (f: A -> B) (a: A) : B = f a
app (A: U) (B: A -> U) (a: A) (f: A -> B a) : B a = f a
\end{lstlisting}
\end{definition}

\begin{theorem} ($\Pi$-Computation).
\begin{lstlisting}
Beta (A: U) (B: A -> U) (a: A) (f: A -> B a)
   : Path (B a) (app A B a (lam A B a (f a))) (f a)
\end{lstlisting}
\end{theorem}

\begin{theorem} ($\Pi$-Uniqueness).
\begin{lstlisting}
Eta (A: U) (B: A -> U) (a: A) (f: A -> B a)
  : Path (A -> B a) f (\(x:A) -> f x)
\end{lstlisting}
\end{theorem}

\subsubsection*{Examples from Mathematics}

The adjoints $\Pi$ and $\Sigma$ is not the only adjoints could be presented in type system.
Axiomatic cohesions could contain a set of adjoint pairs as a core type checker operations.

Geometrically, $\Pi$ type is a space of sections, while the dependent codomain is a space of fibrations.
Lambda functions are sections or points in these spaces, while the function result is a fibration.
$\Pi$ type also represents the cartesian family of sets, generalizing the cartesian product of sets.

\begin{definition} (Section).
A section of morphism $f: A \rightarrow B$ in some category is the morphism $g: B \rightarrow A$
such that $f \circ g: B \mapright{g} A \mapright{f} B$ equals the identity morphism on B.
\end{definition}

\begin{definition} (Fiber).
The fiber of the map $p: E \rightarrow B$ in a point $y: B$ is all points $x: E$ such that $p(x)=y$.
\end{definition}

\begin{definition} (Fiber Bundle).
The fiber bundle $ F \rightarrow E \mapright{p} B$ on a total space $E$ with fiber layer $F$ and base $B$ is a
structure $(F,E,p,B)$ where $p: E \rightarrow B$ is a surjective map with following property:
for any point $y: B$ exists a neighborhood $U_b$ for which a homeomorphism $f: p^{-1}(U_b) \rightarrow U_b \times F$
making the following diagram commute.
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em]
  {
     {p^{-1}(U_b)} & {U_b \times F} \\
     U_b & \\};
  \path[-stealth]
    (m-1-1) edge node [above] {$p$}    (m-1-2)
            edge node [left]  {$f$}    (m-2-1)
    (m-1-2) edge node [right] {$pr_1$} (m-2-1);
\end{tikzpicture}
\end{center}
\end{definition}

\begin{definition} (Cartesian Product of Family over B).
Is a set $F$ of sections of the bundle with elimination map $app : F \times B \rightarrow E$ such that
\begin{equation}
F \times B \mapright{app} E \mapright{pr_1} B
\end{equation}
$pr_1$ is a product projection, so $pr_1$, $app$ are moriphisms
of slice category $Set_{/B}$. The universal mapping property of $F$:
for all $A$ and morphism $A \times B \rightarrow E$ in $Set_{/B}$ exists
unique map $A \rightarrow F$ such that everything commute. So a category
with all dependent products is necessarily a category with all pullbacks.
\end{definition}

\begin{definition} (Trivial Fiber Bundle).
When total space $E$ is cartesian product $\Sigma(B,F)$ and $p = pr_1$
then such bundle is called trivial $(F,\Sigma(B,F),pr_1,B)$.
\end{definition}

\begin{definition} (Dependent Product).
The dependent product along morphism $g: B \rightarrow A$ in category $C$ is the right
adjoint $\Pi_g : C_{/B} \rightarrow C_{/A}$ of the base change functor.
\end{definition}

\begin{definition} (Space of Sections).
Let $\mathbf{H}$ be a $(\infty,1)$-topos, and let $E \rightarrow B : \mathbf{H}_{/B}$ a bundle in
$\mathbf{H}$, object in the slice topos. Then the space of sections $\Gamma_\Sigma(E)$
of this bundle is the Dependent Product:
$$ \Gamma_\Sigma(E) = \Pi_\Sigma (E) \in \mathbf{H}. $$
\end{definition}

\begin{theorem} (Functions Preserve Paths).
For a function $f: (x:A) \rightarrow B(x)$
there is an $ap_f : x =_A y \rightarrow f(x) =_{B(x)} f(y)$. This is called
application of $f$ to path or congruence property (for non-dependent case ---
$cong$ function). This property behaves functoriality
as if paths are groupoid morphisms and types are objects.
\end{theorem}

\begin{theorem} (Trivial Fiber equals Family of Sets).
Inverse image (fiber) of fiber bundle $(F,B*F,pr_1,B)$ in point $y:B$ equals $F(y)$.
\begin{lstlisting}
FiberPi (B: U) (F: B -> U) (y: B)
      : Path U (fiber (Sigma B F) B (pi1 B F) y) (F y)
\end{lstlisting}
\end{theorem}

\begin{theorem} (Homotopy Equivalence).
If fiber space is set for all base, and
there are two functions $f,g : (x:A) \rightarrow B(x)$ and two
homotopies between them, then these homotopies are equal.
\begin{lstlisting}
setPi (A: U) (B: A -> U) (h: (x: A) -> isSet (B x)) (f g: Pi A B)
      (p q: Path (Pi A B) f g) : Path (Path (Pi A B) f g) p q
\end{lstlisting}
\end{theorem}

\begin{theorem} (HomSet).
If codomain is set then space of sections is a set.
\begin{lstlisting}
setFun (A B : U) (_: isSet B) : isSet (A -> B)
\end{lstlisting}
\end{theorem}

\begin{theorem} (Contractability).
If domain and codomain is contractible then the space of sections is contractible.
\begin{lstlisting}
piIsContr (A: U) (B: A -> U) (u: isContr A)
          (q: (x: A) -> isContr (B x)) : isContr (Pi A B)
\end{lstlisting}
\end{theorem}

\newpage
\subsection{Sigma}

$\Sigma$ is a dependent product type, the generalization of products.
$\Sigma$ type is a total space of fibration. Element of total
space is formed as a pair of basepoint and fibration.

\begin{definition} ($\Sigma$-Formation).
\begin{lstlisting}
Sigma (A : U) (B : A -> U) : U = (x : A) * B x
\end{lstlisting}
\end{definition}

\begin{definition} ($\Sigma$-Introduction).
\begin{lstlisting}
dpair (A: U) (B: A -> U) (a: A) (b: B a) : Sigma A B = (a,b)
\end{lstlisting}
\end{definition}

\begin{definition} ($\Sigma$-Elimination).
\begin{lstlisting}
pr1 (A: U) (B: A -> U)
    (x: Sigma A B): A = x.1

pr2 (A: U) (B: A -> U)
    (x: Sigma A B): B (pr1 A B x) = x.2

sigInd (A: U) (B: A -> U) (C: Sigma A B -> U)
       (g: (a: A) (b: B a) -> C (a, b))
       (p: Sigma A B) : C p = g p.1 p.2
\end{lstlisting}
\end{definition}

\begin{theorem} ($\Sigma$-Computation).
\begin{lstlisting}
Beta1 (A: U) (B: A -> U)
      (a:A) (b: B a)
    : Equ A a (pr1 A B (a,b))
    = refl A a

Beta2 (A: U) (B: A -> U)
      (a: A) (b: B a)
    : Equ (B a) b (pr2 A B (a,b))
    = reflect (B a)
\end{lstlisting}
\end{theorem}

\begin{theorem} ($\Sigma$-Uniqueness).
\begin{lstlisting}
Eta2 (A: U) (B: A -> U) (p: Sigma A B)
   : Equ (Sigma A B) p (pr1 A B p,pr2 A B p)
   = refl (Sigma A B) p
\end{lstlisting}
\end{theorem}

\newpage
\subsubsection*{Examples from Mathematics}

\begin{definition} (Dependent Sum).
The dependent sum along the morphism $f: A \rightarrow B$ in category $C$ is the left
adjoint $\Sigma_f : C_{/A} \rightarrow C_{/B}$ of the base change functor.
\end{definition}

\begin{theorem} (Axiom of Choice).
If for all $x : A$ there is $y : B$ such that $R(x,y)$,
then there is a function $f : A \rightarrow B$
such that for all $x : A$ there is a witness of $R(x,f(x))$.
\begin{lstlisting}
ac (A B: U) (R: A -> B -> U)
 : (p: (x:A) -> (y:B)*(R x y)) -> (f:A->B) * ((x:A)->R(x)(f x))
\end{lstlisting}
\end{theorem}

\begin{theorem} (Total).
If fiber over base implies another fiber
over the same base then we can construct total space of section
over that base with another fiber.
\begin{lstlisting}
total (A:U) (B C: A -> U)
      (f: (x:A) -> B x -> C x) (w: Sigma A B)
    : Sigma A C = (w.1,f (w.1) (w.2))
\end{lstlisting}
\end{theorem}

\begin{theorem} ($\Sigma$-Contractability). If the fiber is set then the $\Sigma$ is set.
\begin{lstlisting}
setSig (A:U) (B: A -> U) (sA: isSet A)
       (sB : (x:A) -> isSet (B x)) : isSet (Sigma A B)
\end{lstlisting}
\end{theorem}

\begin{theorem} (Path Between Sigmas).
Path between two sigmas $t,u: \Sigma(A,B)$ could be decomposed to
sigma of two paths $p:t_1=_{A}u_1)$ and $(t_2=_{B(p@i)}u_2)$.
\begin{lstlisting}
pathSig (A:U) (B : A -> U) (t u : Sigma A B)
      : Path U (Path (Sigma A B) t u)
               ((p: Path A t.1 u.1) * PathP (<i>B(p@i)) t.2 u.2)
\end{lstlisting}
\end{theorem}

\newpage
\subsection{Path}

The Path identity type defines a Path space with elements and values.
Elements of that space are functions from interval $[0,1]$ to a values of that path space.
This ctt file reflects \footnote{Cyril Cohen, Thierry Coquand, Simon Huber, Anders M{\"{o}}rtberg. Cubical Type Theory: a constructive interpretation of the univalence axiom. 2015. \url{https://5ht.co/cubicaltt.pdf}}{CCHM} cubicaltt model with connections.
For \footnote{Carlo Angiuli, Brunerie, Coquand, Kuen-Bang Hou (Favonia), Robert Harper, Dan Licata. Cartesian Cubical Type Theory. 2017. \url{https://5ht.co/cctt.pdf}}{ABCFHL} yacctt model with
variables please refer to ytt file. You may also want to
read \footnote{Marc Bezem, Thierry Coquand, Simon Huber. A model of type theory in cubical sets. 2014. \url{http://www.cse.chalmers.se/~coquand/mod1.pdf}}{BCH},
\footnote{Carlo Angiuli, Kuen-Bang Hou (Favonia), Robert Harper. Cartesian Cubical Computational Type Theory: Constructive Reasoning with Paths and Equalities. 2018. \\ \url{https://www.cs.cmu.edu/~cangiuli/papers/ccctt.pdf}}{AFH}.
There is a \footnote{Andrew Pitts, Ian Orton. Axioms for Modelling Cubical Type Theory in a Topos. 2016. \url{https://arxiv.org/pdf/1712.04864.pdf}}{PO} paper about CCHM axiomatic in a topos.

\begin{definition} (Path Formation).
\begin{lstlisting}
Hetero (A B: U) (a: A) (b: B) (P: Path U A B) : U = PathP P a b
Path (A: U) (a b: A) : U = PathP (<i> A) a b
\end{lstlisting}
\end{definition}

\begin{definition} (Path Reflexivity).
Returns an element of reflexivity path space for a given value of the type.
The inhabitant of that path space is the lambda on the homotopy
interval $[0,1]$ that returns a constant value a. Written in
syntax as |<i>a| which equals to $\lambda\ (i: I) \rightarrow a$.
\begin{lstlisting}
refl (A: U) (a: A) : Path A a a
\end{lstlisting}
\end{definition}

\begin{definition} (Path Application).
You can apply face to path.
\begin{lstlisting}
app1 (A: U) (a b: A) (p: Path A a b): A = p @ 0
app2 (A: U) (a b: A) (p: Path A a b): A = p @ 1
\end{lstlisting}
\end{definition}

\begin{definition} (Path Composition).
Composition operation allows to build a new path by given to paths
in a connected point.
\begin{center}
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=3em]
  {
     a & c \\ % (1,1) (1,2)
     a & b \\ % (2,1) (2,2)
  };
  \path[-stealth]
    (m-1-1) edge node [above] {$comp$} (m-1-2)
    (m-2-1) edge node [left]  {$\lambda(i:I)\rightarrow a$} (m-1-1)
    (m-2-2) edge node [right] {$q$} (m-1-2)
    (m-2-1) edge node [above] {$p @ i$} (m-2-2);
\end{tikzpicture}
\end{center}
\begin{lstlisting}
composition (A: U) (a b c: A) (p: Path A a b) (q: Path A b c)
          : Path A a c = comp (<i>Path A a (q@i)) p []
\end{lstlisting}
\end{definition}

\newpage
\begin{theorem} (Path Inversion).
\begin{lstlisting}
inv (A: U) (a b: A) (p: Path A a b): Path A b a = <i> p @ -i
\end{lstlisting}
\end{theorem}

\begin{definition} (Connections).
Connections allows you to build square
with given only one element of path: i) $\lambda\ (i,j: I) \rightarrow p\ @\ min(i,j)$;
ii) $\lambda\ (i,j:I) \rightarrow p\ @\ max(i,j)$.
\begin{center}
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=3em]
  {
     a & b \\ % (1,1) (1,2)
     a & a                    \\ % (2,1) (2,2)
  };
  \path[-stealth]
    (m-1-1) edge node [above] {$p$}    (m-1-2)
    (m-2-1) edge node [left]  {$\lambda\ (i:I)\rightarrow a$}    (m-1-1)
    (m-2-2) edge node [right] {$p$} (m-1-2)
    (m-2-1) edge node [above] {$\lambda\ (i:I)\rightarrow a$} (m-2-2);
  \end{tikzpicture}
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=3em]
  {
     b & b \\ % (1,1) (1,2)
     a & b                    \\ % (2,1) (2,2)
  };
  \path[-stealth]
    (m-1-1) edge node [above] {$\lambda\ (i:I) \rightarrow b$}    (m-1-2)
    (m-2-1) edge node [left]  {$p$}    (m-1-1)
    (m-2-2) edge node [right] {$\lambda\ (i:I) \rightarrow b$} (m-1-2)
    (m-2-1) edge node [above] {$p$} (m-2-2);
  \end{tikzpicture}
\end{center}
\begin{lstlisting}
connection1 (A: U) (a b: A) (p: Path A a b)
          : PathP (<x> Path A (p@x) b) p (<i>b)
          = <y x> p @ (x \/ y)

connection2 (A: U) (a b: A) (p: Path A a b)
          : PathP (<x> Path A a (p@x)) (<i>a) p
          = <x y> p @ (x /\ y)
\end{lstlisting}
\end{definition}

\begin{theorem} (Congruence).
Is a map between values of one type
to path space of another type by an encode function between types.
Implemented as lambda defined on $[0,1]$ that returns
application of encode function to path application of
the given path to lamda argument |$\lambda$ (i:I) $\rightarrow$ f (p @ i)|
for both cases.
\begin{lstlisting}
ap  (A B: U) (f: A -> B)
    (a b: A) (p: Path A a b)
  : Path B (f a) (f b)

apd (A: U) (a x:A) (B: A -> U) (f: A -> B a)
    (b: B a) (p: Path A a x)
  : Path (B a) (f a) (f x)
\end{lstlisting}
\end{theorem}

\begin{theorem} (Transport).
Transports a value of the domain type to the value of the codomain type
by a given path element of the path space between domain and codomain types.
Defined as path composition with |[]| of a over a path $p$ --- |comp p a []|.
\begin{lstlisting}
trans (A B: U) (p: Path U A B) (a: A) : B
\end{lstlisting}
\end{theorem}

\newpage
\begin{definition} (Singleton).
\begin{lstlisting}
singl (A: U) (a: A): U = (x: A) * Path A a x
\end{lstlisting}
\end{definition}

\begin{theorem} (Singleton Instance).
\begin{lstlisting}
eta (A: U) (a: A): singl A a = (a,refl A a)
\end{lstlisting}
\end{theorem}

\begin{theorem} (Singleton Contractability).
\begin{lstlisting}
contr (A: U) (a b: A) (p: Path A a b)
  : Path (singl A a) (eta A a) (b,p)
  = <i> (p @ i,<j> p @ i/\j)
\end{lstlisting}
\end{theorem}


\begin{theorem} (Path Elimination, Diagonal).
\begin{lstlisting}
D (A: U) : U = (x y: A) -> Path A x y -> U
J (A: U) (x y: A) (C: D A)
  (d: C x x (refl A x))
  (p: Path A x y) : C x y p
= subst (singl A x) T (eta A x) (y, p) (contr A x y p) d where
  T (z: singl A x) : U = C x (z.1) (z.2)
\end{lstlisting}
\end{theorem}

\begin{theorem} (Path Elimination, Paulin-Mohring).
J is formulated in a form of Paulin-Mohring and implemented using
two facts that singleton are contractible and dependent function
transport.
\begin{lstlisting}
J (A: U) (a b: A)
  (P: singl A a -> U)
  (u: P (a,refl A a))
  (p: Path A a b) : P (b,p)
\end{lstlisting}
\end{theorem}

\begin{theorem} (Path Elimination, HoTT).
J from HoTT book.
\begin{lstlisting}
J (A: U) (a b: A)
  (C: (x: A) -> Path A a x -> U)
  (d: C a (refl A a))
  (p: Path A a b) : C b p
\end{lstlisting}
\end{theorem}

\begin{theorem} (Path Computation).
\begin{lstlisting}
trans_comp (A: U) (a: A)
  : Path A a (trans A A (<_> A) a)
  = fill (<i> A) a []
subst_comp (A: U) (P: A -> U) (a: A) (e: P a)
  : Path (P a) e (subst A P a a (refl A a) e)
  = trans_comp (P a) e
J_comp (A: U) (a: A) (C: (x: A) -> Path A a x -> U) (d: C a (refl A a))
  : Path (C a (refl A a)) d (J A a C d a (refl A a))
  = subst_comp (singl A a) T (eta A a) d where T (z: singl A a)
  : U = C a (z.1) (z.2)
\end{lstlisting}
\end{theorem}

Note that  Path type has no Eta rule due to groupoid interpretation.

\newpage
\subsection{MLTT}
\begin{definition} (MLTT).
The MLTT as a Type is defined by taking all rules
for $\Pi$, $\Sigma$ and Path type into one $\Sigma$-chain.
\begin{lstlisting}[mathescape=true]
MLTT (A: U): U
  = (Pi_Former: (A -> U) -> U)
  * (Pi_Intro: (B: A -> U) (a: A) -> B a -> (A -> B a))
  * (Pi_Elim: (B: A -> U) (a: A) -> (A -> B a) -> B a)
  * (Pi_Comp1: (B: A -> U) (a: A) (f: A -> B a) ->
    Path (B a) (Pi_Elim B a (Pi_Intro B a (f a))) (f a))
  * (Pi_Comp2: (B: A -> U) (a: A) (f: A -> B a) ->
    Path (A -> B a) f (\(x:A) -> f x))
  * (Sigma_Former: (A -> U) -> U)
  * (Sigma_Intro: (B: A -> U) (a: A) -> (b: B a) -> Sigma A B)
  * (Sigma_Elim1: (B: A -> U) (_: Sigma A B) -> A)
  * (Sigma_Elim2: (B: A -> U) (x: Sigma A B) -> B (pr1 A B x))
  * (Sigma_Comp1: (B: A -> U) (a: A) (b: B a) ->
    Path A a (Sigma_Elim1 B (Sigma_Intro B a b)))
  * (Sigma_Comp2:  (B: A -> U) (a: A) (b: B a) ->
    Path (B a) b (Sigma_Elim2 B (a,b)))
  * (Sigma_Comp3: (B: A -> U) (p: Sigma A B)  ->
    Path (Sigma A B) p (pr1 A B p,pr2 A B p))
  * (Id_Former: A -> A -> U)
  * (Id_Intro: (a: A) -> Path A a a)
  * (Id_Elim: (x: A) (C: D A) (d: C x x (Id_Intro x))
    (y: A) (p: Path A x y) -> C x y p)
  * (Id_Comp: (a:A)(C: D A) (d: C a a (Id_Intro a)) ->
    Path (C a a (Id_Intro a)) d (Id_Elim a C d a (Id_Intro a))) * U
\end{lstlisting}
\end{definition}

\begin{theorem} (Sound Check).
There is an instance of MLTT.
\begin{lstlisting}
instance (A: U): MLTT A
    = (Pi A, lam A, app A, Beta A, Eta A,
       Sigma A, dpair A, pr1 A, pr2 A, Beta1 A, Beta2 A, Eta2 A,
       Path A, refl A, J A, J_comp A, A)
\end{lstlisting}
\end{theorem}

\subsection*{MLTT Model Check}

The result of the work is a |mltt.ctt| file which can be runned using |cubicaltt|.
Note that computation rules take a seconds to type check.

\begin{lstlisting}
$ time cubical -b mltt.ctt
Checking: MLTT
Checking: instance
File loaded.

real    0m6.308s
user    0m6.278s
sys     0m0.014s
\end{lstlisting}

\section{Inductive Types}
\subsection{Empty}
\subsection{Unit}
\subsection{Bool}
\subsection{Either}
\subsection{Nat}
\subsection{List}
\subsection{Fin}
\subsection{Vector}

\section{Homotopy Type Theory}
\subsection{Functional Extensionality}
\begin{definition} (Formation).
\end{definition}

\begin{theorem} (Introduction).
\end{theorem}

\begin{theorem} (Elimination).
\end{theorem}

\begin{theorem} (Computation).
\end{theorem}

\begin{theorem} (Uniqueness).
\end{theorem}
\subsection{n-Types}
\subsection{Loop Space}
\subsection{Pointed Types}

\subsection{Univalence}

\begin{definition} (Formation).
\end{definition}

\begin{theorem} (Introduction).
\end{theorem}

\begin{theorem} (Elimination).
\end{theorem}

\begin{theorem} (Computation).
\end{theorem}

\begin{theorem} (Uniqueness).
\end{theorem}

\subsection{Homotopy Group}

\section{Higher Inductive Types}

CW-complexes are fundamental objects in homotopy type theory
and even included inside cubical type checker in a form of
higher (co)-inductive types (HITs).
Just like regular (co)-inductive types could be described as recursive
terminating (well-founded) or non-terminating trees,
higher inductive types could be described as CW-complexes.
Defining HIT means to define some CW-complex
directly using cubical homogeneous composition structure as an
element of initial algebra inside cubical model.

\subsection{Interval}
\subsection{Pullback}
\subsection{Pushout}
\subsection{Suspension}
\subsection{Truncation}
\subsection{Quotient}
\subsection{Hopf Fibration}

\section{Cagegory Theory}
\subsection{Category}
\subsection{Functor}
\subsection{Adjunction}
\subsection{Natural Transformation}
\subsection{Pullback}
\subsection{Slice}
\subsection{Terminals}

\section{Topos Theory}
\subsection{Topology}
\subsection{Coverage}
\subsection{Grothendieck Topology}
\subsection{Grothendieck Topos}
\subsection{Elementary Topos}
\subsection{Geometric Morphism}
\subsection{Category of Topoi}

\section{Cohesive Type Theory}
\subsection{Fundametal $\infty$-Groupoid}
\subsection{Infinitesimal Modality}
\subsection{Sharp Modality}
\subsection{Formal Disc}
\subsection{Etale Maps}
\subsection{Manifolds}

%\newpage
\bibliographystyle{plain}
\bibliography{types}

\end{document}

